%% I'd like to thank the academy, but I cannot in good conscience.

\begin{centering}
\textbf{ACKNOWLEDGEMENTS}\\
\vspace{\baselineskip}
\end{centering}

A doctoral thesis is a heck of a thing, and I could not have done it without
the help of a great number of people, some of whom I would like to acknowledge
here.

I'd first like to thank Lindsey Kuper, my lovely wife, for putting up with me
(in general, but especially) throughout our tenure in graduate school. We're
out of the cold now.

I'd like to thank my wonderful parents, Lenoir and Jim, for their continuous
love, encouragement and confidence.
I'd like to thank some of my dearest lifelong friends for their support and
inspiration, both recently and throughout our lives.
In addition to his perennial warmth, humor and curiosity, Ryan Allen gave me an
enormous boost by completing his own doctoral work\cite{allen2017visions}; he
powered through some occasionally less-than-ideal circumstances to finish what
he set out to do. Martin Robinson, Brett Thompson and Sydney Jackson (among
others!) listened to me complain and shared in my joys, as they've been known
to do, providing sounding boards for ideas, both sensible and absurd, and
emotional support as necessary.

I'd like to thank the Google Atlanta diaspora, especially Joel Webber and Bruce
Johnson, for helping me get a foot in the door at one of the world's great
machine learning and NLP companies. You did so much for my confidence and
drive, and you sent me off to graduate school with a spring in my step and a
chip on my shoulder. I hope you realize how much it's meant to me. I'd also
like to thank members of the Google Translate team, both current and former,
for being such great mentors, colleagues and friends. Apurva Shah, Jeff
Klinger, Klaus Macherey, Wolfgang Macherey, Jason Smith, Lee Marshall, Julie
Cattiau, Richard Zens, Mengmeng Niu, Manisha Jain and Macduff Hughes all come
to mind as especially helpful and supportive, but the whole Google Translate
crew is lovely people, and if you happen to be reading this, you probably
helped me out, so thank you.

I'd like to thank the Bloomington Area Runners Association, especially Katie,
Andrew, Zach, Cliff, Mike, Sara, Sarah, and Ben \& Steph, for your friendship
and camaraderie, keeping me relatively healthy and happy during my time in
Bloomington.

And I'd like to thank my committee for their encouragement, support, warmth,
skepticism, feedback, and patience.
